---
title: "Model Fitting"
date: "4/4/2019"
output:
  html_document:
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(manipulate)# Need for manipulate 
library(FME)# Needed for modFit function
library(tidyverse)# Needed for plotting and data manipulation
library(readxl) # Needed for reading in data. 
library(lubridate) # Needed for date and time data manipulation
cbPalette <- c("#009E73", "#0072B2", "#999999", "#E69F00", "#56B4E9",  "#F0E442",  "#D55E00")

```


# Model Fitting 

## Lotka Voltera Comeptition

In this document we will be fitting the Lotka Volterra Competition model (as seen below). 

$$\frac{dS}{dt}=r_S*S(1-\frac{S}{K_S})-\alpha_S*S*C$$

$$\frac{dC}{dt}=r_C*C(1-\frac{C}{K_C})-\alpha_C*S*C$$


In this chunk we read in the data, created two new columns that are the average population count on each day, and then created a day column that is from day 1 to dat 19. Note, for this chunck to work you need to add the file path to where the data is located on your computer.

```{r Reading in Data}
algae_dilluted_data<- read_excel("~/MathBio/Data/Algae Data Diluted.xlsx")# Reading in the data
start_date=algae_dilluted_data$Date[1]# Grabbing what the first day we took data is. 
algae_dilluted_data=algae_dilluted_data%>%
  mutate(Scene_Average=(Scene_Count_1+Scene_Count_2+Scene_Count_3+Scene_Count_4# Calculating the average scene on each day
                        +Scene_Count_5+Scene_Count_6+Scene_Count_7+Scene_Count_8)/8,
         Clamy_Average=(Clamy_Count_1+Clamy_Count_2+Clamy_Count_3+Clamy_Count_4# Calculating the average chlamy on each day. 
                        +Clamy_Count_5+Clamy_Count_6+Clamy_Count_7+Clamy_Count_8)/8)%>%
  mutate(day=as.numeric(difftime(Date, start_date, units="days"))+1)

algae_dilluted_data_subset=algae_dilluted_data%>%
  select(day, Flask, Scene_Average, Clamy_Average) # selecting only the columns that I will need going forward. 
```

Here we created our function in r to describe the differential equations and then tested it with some parameter estimates and plotted it to make sure everything is working right.

```{r model building and testing}
compderivs=function(t,pop,parms) { # function to describe differential equations
  with(as.list(c(pop,parms)), {
    S=pop[1]
    C=pop[2]
    rS=parms[1]
    kS=parms[2]
    alphaS=parms[3]
    rC=parms[4]
    kC=parms[5]
    alphaC=parms[6]
    dS=rS*S*(1-(S/kS))-alphaS*S*C
    dC=rC*C*(1-(C/kC))-alphaC*C*S
    list(c(dS,dC))
  }
)}


yinitcomp=c(0.01,0.01) #initial conditions
parms=c(.0001,400,.004,.07,.400,.04) # parameter esitmates (really just some random numbers we picked)
t=seq(0,19, length=1000) # our time
outcomp=ode(y=yinitcomp,times=t,func = compderivs, parms=parms) # numerically solving thwe differential equation


outcomp=as_tibble(outcomp) # Making a tibble so that we can plot it easier
comp_test=outcomp%>% # renaming the columns
  rename("Scene"=`1`, 
         "Clamy"=`2`)

comp_test%>% # plotting just to test everything out
  ggplot()+
  geom_line(aes(x=time, y=Scene), color="red")+
  geom_line(aes(x=time, y=Clamy), color="blue")

```

## Manipulate General Usage

This is the code that we used for finding parameter estimates for all of our models. When we switch flasks we just change the points that we are plotting to the corresponding flask. Note, for this to work you need to add the file path to the data on your computer. 

```{r, eval=FALSE}
algae_dilluted_data<- read_excel("~/MathBio/Data/Algae Data Diluted.xlsx") # loading data
start_date=algae_dilluted_data$Date[1]
algae_dilluted_data=algae_dilluted_data%>% # cleaning up the data and taking some averages
  mutate(Scene_Average=(Scene_Count_1+Scene_Count_2+Scene_Count_3+Scene_Count_4
                        +Scene_Count_5+Scene_Count_6+Scene_Count_7+Scene_Count_8)/8,
         Clamy_Average=(Clamy_Count_1+Clamy_Count_2+Clamy_Count_3+Clamy_Count_4
                        +Clamy_Count_5+Clamy_Count_6+Clamy_Count_7+Clamy_Count_8)/8)%>%
  mutate(day=as.numeric(difftime(Date, start_date, units="days"))+1)

algae_dilluted_data_1=algae_dilluted_data%>%
  filter(Flask=="1.7DC")%>%# selecting the flask we want to look at 
  select(Scene_Average, Clamy_Average, day)


algae_dilluted_data_subset=algae_dilluted_data%>% # selecting only the columns that we want
  select(day, Flask, Scene_Average, Clamy_Average)

algae_dilluted_data_1.7DC_S=algae_dilluted_data_1%>% # making a scene dataset
  select(Scene_Average, day)

algae_dilluted_data_1.7DC_C=algae_dilluted_data_1%>% # making a chlamy dataset
  select(Clamy_Average, day)


manipulate({
  compderivs=function(t,pop,parms) { # function to descrive model 
    with(as.list(c(pop,parms)), {
      S=pop[1]
      C=pop[2]
      rS=parms[1]
      kS=parms[2]
      alphaS=parms[3]
      rC=parms[4]
      kC=parms[5]
      alphaC=parms[6]
      dS=rS*S*(1-(S/kS))-alphaS*S
      dC=rC*C*(1-(C/kC))-alphaC*C
      list(c(dS,dC))
    }
    )}
  
  
  yinitcomp=c(0.01,0.01) # initial conditions
  parms=c(rSs,KSs,alphaSs,rCs,KCs,alphaCs) # parameter values (they are this way for the sliders)
  t=seq(0,19, length=1000) # time that we are looking at
  outcomp=ode(y=yinitcomp,times=t,func = compderivs, parms=parms) # solve with the parameters that we currently have
  outcomp=as_tibble(outcomp) # Make it into a tibble to make the plotting easier. 
  comp_test=outcomp%>%
    rename("Scene"=`1`, # rename the columns 
           "Clamy"=`2`) 
  
 plot(algae_dilluted_data_1$day, algae_dilluted_data_1$Scene_Average, col="red",  xlim=c(0,20), ylim=c(0,30), xlab="Days", ylab="Number of Cells *10^4 cells/ml", main="Manipulate For Parameter Esitmates Plot")# plotting scene
 points(algae_dilluted_data_1$day, algae_dilluted_data_1$Clamy_Average, col="blue", xlim=c(0,20)) # plotting chlamy
 lines(comp_test$time, comp_test$Scene, col="red") # plotting scene prediction
 lines(comp_test$time, comp_test$Clamy, col="blue")# plotting chlamy prediction
 legend("topright",col = c("blue", "red"), legend=c("Chlamy", "Scene"), pch=c(1,1), lty=c(1,1), cex=0.5) # Adding a legend 
 
  
}, rSs=slider(0,2,step=0.001, initial=1.3), #sliders so that we can change the params need one slider for each param.  
KSs=slider(0,100, initial =5, step=0.1),
alphaSs=slider(0,2,step=0.001, initial = 0.029),
rCs=slider(0,1.5,step=0.001, initial=.763),
KCs=slider(0,100, initial = 90, step=0.1),
alphaCs=slider(0,1,step=0.001, initial = 0.07))
```



## 1.7DC

Here we fit the model to the 1.7mM diluted and combined sample and plotted the results. 

```{r fitting to 1.7DC}
algae_dilluted_data_17dc=algae_dilluted_data%>% 
  filter(Flask=="1.7DC")%>% # getting only the flask we want.
  select(Scene_Average, Clamy_Average, day, Flask)%>%
  mutate(Scene_Average=ifelse(day==19, 4.62, Scene_Average), # We accidentally entered the last two points into the dataframe wrong so we are just changing that here  
         Clamy_Average=ifelse(day==19, 23.8, Clamy_Average))


algae_dilluted_data_1.7DC_S=algae_dilluted_data_17dc%>% # making a scene dataset
  select(Scene_Average, day)
algae_dilluted_data_1.7DC_C=algae_dilluted_data_17dc%>% # making a chlamy dataset
  select(Clamy_Average, day)

tcomp=unique(algae_dilluted_data$day) #changing our time so that it is the same as when we have data points. 

sse.comp=function(parms0){  #function to calculate squared errors
    rS=parms0[1]
    kS=parms0[2]
    alphaS=parms0[3]
    rC=parms0[4]
    kC=parms0[5]
    alphaC=parms0[6]
  out1comp=ode(y = yinitcomp, times = tcomp, func = compderivs, parms=c(rS,kS,alphaS,rC,kC,alphaC)) # numerically solve differential equations
  return(c(((out1comp[,2]-algae_dilluted_data_1.7DC_S$Scene_Average)^2),(out1comp[,3]-algae_dilluted_data_1.7DC_C$Clamy_Average)^2))   #squared errors
}


parms17dc=c(1,5,0.029,1.5,20,0.07)# These were found using mainpulate

fitcomp_17dc=modFit(f=sse.comp, p=parms17dc, lower=c(0,0,0,0,0,0), upper=c(5,100,2,2,50,2)) # modfit to minimize the sum of squared error. 

fitcomp_17dc$par # take a peek at our parameter estimates 

out2comp_17dc=ode(y=yinitcomp, times=t, func = compderivs, parms=fitcomp_17dc$par) # numerically solve wiht the optimal parameter estimates

dc17_plot=out2comp_17dc%>% # plot the results
  as.matrix()%>%
  as_tibble()%>%
  rename("Scene"=`1`)%>%
  rename("Clamy"=`2`)%>%
  ggplot()+
  geom_line(aes(x=time, y=Clamy, color="#009E73"))+
  geom_line(aes(x=time, y=Scene, color="#0072B2"))+
  geom_point(data=algae_dilluted_data_1.7DC_C, aes(x=day, y=Clamy_Average), color="#009E73")+
  geom_point(data=algae_dilluted_data_1.7DC_S, aes(x=day, y=Scene_Average), color="#0072B2")+
  ggtitle("1.7DC Lotka Volterra")+
  scale_colour_manual(name = 'Population', values =c("#009E73", "#0072B2"), labels = c("Chlamydomonas","Scenedesmus"))+
  xlab("Days")+
  ylab("Number of Cells *10^4 per ml")+
  theme_light()

dc17_plot

```



##0.34DC

Here we fit the model to the 0.34 mM diluted and combined data. 

```{r}
algae_dilluted_data_34dc=algae_dilluted_data%>%
  filter(Flask=="0.34DC")%>% # getting only the flask we want. 
  select(Scene_Average, Clamy_Average, day)

algae_dilluted_data_34dc_S=algae_dilluted_data_34dc%>%# Make one data set for each species.
  select(Scene_Average, day)
algae_dilluted_data_34dc_C=algae_dilluted_data_34dc%>%
  select(Clamy_Average, day)


tcomp=unique(algae_dilluted_data_34dc$day) # make sure time is the same as when we have data 

sse.comp=function(parms0){  #function to calculate squared errors
    rS=parms0[1]
    kS=parms0[2]
    alphaS=parms0[3]
    rC=parms0[4]
    kC=parms0[5]
    alphaC=parms0[6]
  out1comp=ode(y = yinitcomp, times = tcomp, func = compderivs, parms=c(rS,kS,alphaS,rC,kC,alphaC))
  return(c(((out1comp[,2]-algae_dilluted_data_34dc_S$Scene_Average)^2),(out1comp[,3]-algae_dilluted_data_34dc_C$Clamy_Average)^2))   #sum of squared errors
}
parms_34dc=c(.536,25,0.029,0.87,8.5,0.04) # initial guesses from manipulate. 
fitcomp_34dc=modFit(f=sse.comp, p=parms_34dc,lower=c(0,0,0,0,0,0),upper=c(2,100,2,2,15,2))# modFit to minimize SSE

fitcomp_34dc$par # look at the parameter estimates

out2comp_34dc=ode(y=yinitcomp,times=t,func = compderivs,parms=fitcomp_34dc$par)# numerically solve with the optimal parameter estimates. 

dc34_plot=out2comp_34dc%>%# plot the results. 
  as_tibble()%>%
  rename("Scene"=`1`)%>%
  rename("Clamy"=`2`)%>%
  ggplot()+
  geom_line(aes(x=time, y=Clamy, color="red"))+
  geom_line(aes(x=time, y=Scene, color="blue"))+
  geom_point(data=algae_dilluted_data_34dc_C, aes(x=day, y=Clamy_Average, color="red"))+
  geom_point(data=algae_dilluted_data_34dc_S, aes(x=day, y=Scene_Average, color="blue"))+
  ggtitle("0.34DC Lotka Volterra")+
  scale_colour_manual(name = 'Population', values =c( "#0072B2", "#009E73"), labels = c("Scenedesmus","Chlamydomonas"))+
  xlab("Days")+
  ylab("Number of Cells *10^4 per ml")+
  theme_light()
dc34_plot
```

## 5DC

Here we fit our model to our 5$\mu$M diluted and combined data. The procedure is the same as the first two sets of data that we fit. 

```{r}
algae_dilluted_data_5dc=algae_dilluted_data%>%
  filter(Flask=="5DC")%>%
  select(Scene_Average, Clamy_Average, day)

algae_dilluted_data_5dc_S=algae_dilluted_data_5dc%>%
  select(Scene_Average, day)
algae_dilluted_data_5dc_C=algae_dilluted_data_5dc%>%
  select(Clamy_Average, day)

sse.comp=function(parms0){  #function to calculate squared errors
    rS=parms0[1]
    kS=parms0[2]
    alphaS=parms0[3]
    rC=parms0[4]
    kC=parms0[5]
    alphaC=parms0[6]
  out1comp=ode(y = yinitcomp, times = tcomp, func = compderivs, parms=c(rS,kS,alphaS,rC,kC,alphaC))
  return(c(((out1comp[,2]-algae_dilluted_data_5dc_S$Scene_Average)^2),(out1comp[,3]-algae_dilluted_data_5dc_C$Clamy_Average)^2))   #sum of squared errors
}
parms_5dc=c(1.178,19.3,.979,.555,5,.0846)
fitcomp_5dc=modFit(f=sse.comp, p=parms_5dc,lower=c(1,18,0,0,0,0),upper=c(2,30,1,1,10,.4))
fitcomp_5dc$par
out2comp_5dc=ode(y=yinitcomp,times=t,func = compderivs,parms=fitcomp_5dc$par)

dc5_plot=out2comp_5dc%>%
  as_tibble()%>%
  rename("Scene"=`1`)%>%
  rename("Clamy"=`2`)%>%
  ggplot()+
  geom_line(aes(x=time, y=Clamy, color="red"))+
  geom_line(aes(x=time, y=Scene, color="blue"))+
  geom_point(data=algae_dilluted_data_5dc_C, aes(x=day, y=Clamy_Average, color="red"))+
  geom_point(data=algae_dilluted_data_5dc_S, aes(x=day, y=Scene_Average, color="blue"))+
  ggtitle("5DC Lotka Volterra")+
  scale_colour_manual(name = 'Population', values =c( "#0072B2", "#009E73"), labels = c("Scenedesmus","Chlamydomonas"))+
  xlab("Days")+
  ylab("Number of Cells *10^4 per ml")+
  theme_light()
dc5_plot
```

## Nice Plot 

Here we made the nice plot that you see on our poster. 

```{r}
out2comp_17dc=as.data.frame(out2comp_17dc) # getting all the data 
out2comp_17dc$Flask="1.7DC"
out2comp_34dc=as.data.frame(out2comp_34dc)
out2comp_34dc$Flask="0.34DC"
out2comp_5dc=as.data.frame(out2comp_5dc)
out2comp_5dc$Flask="5DC"
cbPalette <- c("#E69F00", "#0072B2", "#999999", "#E69F00", "#56B4E9",  "#F0E442", "#009E73" ) # selecting the color palette


clamy_sd <- algae_dilluted_data%>% # reshaping the data
  select(11:18)%>%
  apply(1, sd)%>%
  tibble() %>%
  rename("clamy_sd" = ".")


scene_sd <- algae_dilluted_data%>% # reshaping the data
  select(3:10)%>%
  apply(1, sd)%>%
  tibble()%>%
  rename("scene_sd" = ".")

algae_dilluted_data <- bind_cols(algae_dilluted_data, clamy_sd, scene_sd) # joining them together



dc_plotting_actual=algae_dilluted_data%>% # getting the data ready (more reshaping and cleaning)
  filter(Flask=="0.34DC"| Flask=="1.7DC"| Flask=="5DC")%>%
  select(Scene_Average, Clamy_Average, scene_sd, clamy_sd, day, Flask)%>%
  gather("Species", "Count", 1:2)%>%
  separate(sep="_", Species, into=c("Species", "Trash"))%>%
  select(-Trash)%>%
  mutate(Species=ifelse(Species=="Scene", "Scenedesmus", "Chlamydomonas"))%>%
  mutate(Flask = ifelse(Flask == "1.7DC", "1.7 mM Diluted and Combined", Flask), 
         Flask = ifelse(Flask == "0.34DC", "0.34 mM Diluted and Combined", Flask),
         Flask = ifelse(Flask == "5DC", "5 uM Diluted and Combined", Flask)) %>%
  mutate(Flask = factor((Flask), levels = c("5 uM Diluted and Combined", "0.34 mM Diluted and Combined", "1.7 mM Diluted and Combined")))




dc_plot_data=out2comp_17dc%>% # getting the model fits ready (more reshaping and cleanign)
  rbind(out2comp_34dc)%>%
  rbind(out2comp_5dc)%>%
  as_tibble()%>%
  rename("Scenedesmus"=`1`,
         "Chlamydomonas"=`2`)%>%
  gather("Species", "Count", 2:3)%>%
  mutate(Flask = ifelse(Flask == "1.7DC", "1.7 mM Diluted and Combined", Flask), 
         Flask = ifelse(Flask == "0.34DC", "0.34 mM Diluted and Combined", Flask),
         Flask = ifelse(Flask == "5DC", "5 uM Diluted and Combined", Flask)) %>%
  mutate(Flask = factor((Flask), levels = c("5 uM Diluted and Combined", "0.34 mM Diluted and Combined", "1.7 mM Diluted and Combined")))


ggplot()+ # plotting
  geom_errorbar(data=dc_plotting_actual, # adding error bars to all the points
                aes(x= day, 
                    ymin=Count-clamy_sd/sqrt(8),
                    ymax=Count+clamy_sd/sqrt(8)), 
                  colour="grey",
                  width=.5)+
  geom_point(data=dc_plotting_actual, aes(x=day, y=Count, color=Species), size = 2 )+ # adding the points
  geom_line(data=dc_plot_data, aes(x=time, y=Count, color=Species), size  = 2)+ # adding the best fit lines. 
  facet_wrap(~Flask)+
  ylab("Average Count (x10^4 cells/ml)")+
  xlab("Days")+
  theme_minimal()+
  theme(legend.position="bottom",
        text = element_text(size = 20),
        axis.line = element_line(color="black"))+
  theme(plot.margin=unit(c(1,1,1.5,1.2),"cm"))+
  scale_colour_manual(values=cbPalette)
  
```

## Parameter estimates

Here we compare all of the parameter estimates for the the diluted and combined data. 

```{r}
fitcomp_17dc$par
fitcomp_34dc$par
fitcomp_5dc$par
```


## 1.7D

Same procedure as the diluted and combined. 

```{r}
algae_dilluted_data_17d=algae_dilluted_data%>%
  filter(Flask=="1.7D")%>%
  select(Scene_Average, Clamy_Average, day)

algae_dilluted_data_1.7d_S=algae_dilluted_data_17d%>%
  select(Scene_Average, day)
algae_dilluted_data_1.7d_C=algae_dilluted_data_17d%>%
  select(Clamy_Average, day)
sse.comp=function(parms0){  #function to calculate squared errors
    rS=parms0[1]
    kS=parms0[2]
    alphaS=parms0[3]
    rC=parms0[4]
    kC=parms0[5]
    alphaC=parms0[6]
  out1comp=ode(y = yinitcomp, times = tcomp, func = compderivs, parms=c(rS,kS,alphaS,rC,kC,alphaC))
  return(c(((out1comp[,2]-algae_dilluted_data_1.7d_S$Scene_Average)^2),(out1comp[,3]-algae_dilluted_data_1.7d_C$Clamy_Average)^2))   #sum of squared errors
}


parms_17d=c(.443,4,.041,.873,25.9,.43)
fitcomp_17d=modFit(f=sse.comp, p=parms_17d,lower=c(0,0,0,0,0,0),upper=c(2,10,2,2,100,2))

out2comp_17d=ode(y=yinitcomp,times=t,func = compderivs,parms=fitcomp_17d$par)

out2comp_17d%>%
  as_tibble()%>%
  rename("Scene"=`1`)%>%
  rename("Clamy"=`2`)%>%
  ggplot()+
  geom_line(aes(x=time, y=Clamy, color="red"))+
  geom_line(aes(x=time, y=Scene, color="blue"))+
  geom_point(data=algae_dilluted_data_1.7d_C, aes(x=day, y=Clamy_Average, color="red"))+
  geom_point(data=algae_dilluted_data_1.7d_S, aes(x=day, y=Scene_Average, color="blue"))+
  ggtitle("1.7D Lotka Volterra")+
  scale_colour_manual(name = 'Population', values =c( "#0072B2", "#009E73"), labels = c("Chlamydomonas","Scenedesmus"))+
  xlab("Days")+
  ylab("Number of Cells *10^4 per ml")+
  theme_light()

```



## 0.34D

Same procedure as the diluted and combined. 

```{r}
algae_dilluted_data_34d=algae_dilluted_data%>%
  filter(Flask=="0.34D")%>%
  select(Scene_Average, Clamy_Average, day)
algae_dilluted_data_0.34d_S=algae_dilluted_data_34d%>%
  select(Scene_Average, day)
algae_dilluted_data_0.34d_C=algae_dilluted_data_34d%>%
  select(Clamy_Average, day)

sse.comp=function(parms0){  #function to calculate squared errors
    rS=parms0[1]
    kS=parms0[2]
    alphaS=parms0[3]
    rC=parms0[4]
    kC=parms0[5]
    alphaC=parms0[6]
  out1comp=ode(y = yinitcomp, times = tcomp, func = compderivs, parms=c(rS,kS,alphaS,rC,kC,alphaC))
  return(c(((out1comp[,2]-algae_dilluted_data_0.34d_S$Scene_Average)^2),(out1comp[,3]-algae_dilluted_data_0.34d_C$Clamy_Average)^2))   #sum of squared errors
}


parms_34d=c(0.9,19,0.621,.785,5,.342)


fitcomp_34d=modFit(f=sse.comp,p=parms_34d,lower=c(0,0,0,0,0,0),upper=c(2,20,2,2,150,2))
fitcomp_34d$par

out2comp_34d=ode(y=yinitcomp,times=t,func = compderivs,parms=fitcomp_34d$par)
out2comp_34d%>%
  as_tibble()%>%
  rename("Scene"=`1`)%>%
  rename("Clamy"=`2`)%>%
  ggplot()+
  geom_line(aes(x=time, y=Clamy, color="red"))+
  geom_line(aes(x=time, y=Scene, color="blue"))+
  geom_point(data=algae_dilluted_data_0.34d_C, aes(x=day, y=Clamy_Average, color="red"))+
  geom_point(data=algae_dilluted_data_0.34d_S, aes(x=day, y=Scene_Average, color="blue"))+
  ggtitle("0.34D Lotka Volterra")+
  scale_colour_manual(name = 'Population', values =c( "#0072B2", "#009E73"), labels = c("Scenedesmus","Chlamydomonas"))+
  xlab("Days")+
  ylab("Number of Cells *10^4 per ml")+
  theme_light()

```


## 5D

Same procedure as the diluted and combined. 

```{r}
algae_dilluted_data_5d=algae_dilluted_data%>%
  filter(Flask=="5D")%>%
  select(Scene_Average, Clamy_Average, day)

algae_dilluted_data_5d_S=algae_dilluted_data_5d%>%
  select(Scene_Average, day)
algae_dilluted_data_5d_C=algae_dilluted_data_5d%>%
  select(Clamy_Average, day)


sse.comp=function(parms0){  #function to calculate squared errors
    rS=parms0[1]
    kS=parms0[2]
    alphaS=parms0[3]
    rC=parms0[4]
    kC=parms0[5]
    alphaC=parms0[6]
  out1comp=ode(y = yinitcomp, times = tcomp, func = compderivs, parms=c(rS,kS,alphaS,rC,kC,alphaC))
  return(c(((out1comp[,2]-algae_dilluted_data_5d_S$Scene_Average)^2),(out1comp[,3]-algae_dilluted_data_5d_C$Clamy_Average)^2))   #sum of squared errors
}


parms_5d=c(.536,25,0.029,0.87,8.5,0.04)
fitcomp_5d=modFit(f=sse.comp, p=parms_5d,lower=c(0,0,0,0,0,0),upper=c(2,100,2,2,15,2))
fitcomp_5d$par
out2comp_5d=ode(y=yinitcomp,times=t,func = compderivs,parms=fitcomp_5d$par)

out2comp_5d%>%
  as_tibble()%>%
  rename("Scene"=`1`)%>%
  rename("Clamy"=`2`)%>%
  ggplot()+
  geom_line(aes(x=time, y=Clamy, color="red"))+
  geom_line(aes(x=time, y=Scene, color="blue"))+
  geom_point(data=algae_dilluted_data_5d_C, aes(x=day, y=Clamy_Average, color="red"))+
  geom_point(data=algae_dilluted_data_5d_S, aes(x=day, y=Scene_Average, color="blue"))+
  ggtitle("5D Lotka Volterra")+
  scale_colour_manual(name = 'Population', values =c( "#0072B2", "#009E73"), labels = c("Scenedesmus","Chlamydomonas"))+
  xlab("Days")+
  ylab("Number of Cells *10^4 per ml")+
  theme_light()
```


## 1.7to5D

Same procedure as the diluted and combined. 

```{r}
algae_dilluted_data_17to5d=algae_dilluted_data%>%
  filter(Flask=="1.7to5")%>%
  select(Scene_Average, Clamy_Average, day)

algae_dilluted_data_1.7to5d_S=algae_dilluted_data_17to5d%>%
  select(Scene_Average, day)
algae_dilluted_data_1.7to5d_C=algae_dilluted_data_17to5d%>%
  select(Clamy_Average, day)


sse.comp=function(parms0){  #function to calculate squared errors
    rS=parms0[1]
    kS=parms0[2]
    alphaS=parms0[3]
    rC=parms0[4]
    kC=parms0[5]
    alphaC=parms0[6]
  out1comp=ode(y = yinitcomp, times = tcomp, func = compderivs, parms=c(rS,kS,alphaS,rC,kC,alphaC))
  return(c(((out1comp[,2]-algae_dilluted_data_1.7to5d_S$Scene_Average)^2),(out1comp[,3]-algae_dilluted_data_1.7to5d_C$Clamy_Average)^2))   #sum of squared errors
}

parms_17to5d=c(1.4,27,.1,1.6,11,.1)
fitcomp_17to5d=modFit(f=sse.comp,
                      p=parms_17to5d,lower=c(0,0,0,0,0,0),upper=c(2,100,2,2,15,2))
fitcomp_17to5d$par
out2comp_17to5d=ode(y=yinitcomp,times=t,func = compderivs,parms=fitcomp_17to5d$par)

out2comp_17to5d%>%
  as_tibble()%>%
  rename("Scene"=`1`)%>%
  rename("Clamy"=`2`)%>%
  ggplot()+
  geom_line(aes(x=time, y=Clamy, color="red"))+
  geom_line(aes(x=time, y=Scene, color="blue"))+
  geom_point(data=algae_dilluted_data_1.7to5d_C, aes(x=day, y=Clamy_Average, color="red"))+
  geom_point(data=algae_dilluted_data_1.7to5d_S, aes(x=day, y=Scene_Average, color="blue"))+
  ggtitle("1.7to5D Lotka Volterra")+
  scale_colour_manual(name = 'Population', values =c( "#0072B2", "#009E73"), labels = c("Scenedesmus","Chlamydomonas"))+
  xlab("Days")+
  ylab("Number of Cells *10^4 per ml")+
  theme_light()
```


## 5to1.7

Same procedure as the diluted and combined. 

```{r}
algae_dilluted_data_5to1.7d=algae_dilluted_data%>%
  filter(Flask=="5t01.7")%>%
  select(Scene_Average, Clamy_Average, day)
algae_dilluted_data_5to1.7d
algae_dilluted_data_5to1.7d_S=algae_dilluted_data_5to1.7d%>%
  select(Scene_Average, day)
algae_dilluted_data_5to1.7d_C=algae_dilluted_data_5to1.7d%>%
  select(Clamy_Average, day)

sse.comp=function(parms0){  #function to calculate squared errors
    rS=parms0[1]
    kS=parms0[2]
    alphaS=parms0[3]
    rC=parms0[4]
    kC=parms0[5]
    alphaC=parms0[6]
  out1comp=ode(y = yinitcomp, times = tcomp, func = compderivs, parms=c(rS,kS,alphaS,rC,kC,alphaC))
  return(c(((out1comp[,2]-algae_dilluted_data_5to1.7d_S$Scene_Average)^2),(out1comp[,3]-algae_dilluted_data_5to1.7d_C$Clamy_Average)^2))   #sum of squared errors
}
algae_dilluted_data_5to1.7d

parms_5to1.7d=c(0.9,12,.3,.8,5,.1)
fitcomp_5to1.7d=modFit(f=sse.comp, p=parms_5to1.7d,lower=c(0,0,0,0,0,0),upper=c(2,150,2,2,15,2))
fitcomp_5to1.7d$par
out2comp_5to1.7d=ode(y=yinitcomp,times=t,func = compderivs,parms=fitcomp_5to1.7d$par)

out2comp_5to1.7d%>%
  as_tibble()%>%
  rename("Scene"=`1`)%>%
  rename("Clamy"=`2`)%>%
  ggplot()+
  geom_line(aes(x=time, y=Clamy, color="red"))+
  geom_line(aes(x=time, y=Scene, color="blue"))+
  geom_point(data=algae_dilluted_data_5to1.7d_C, aes(x=day, y=Clamy_Average, color="red"))+
  geom_point(data=algae_dilluted_data_5to1.7d_S, aes(x=day, y=Scene_Average, color="blue"))+
  ggtitle("5to1.7d Lotka Volterra")+
  scale_colour_manual(name = 'Population', values =c( "#0072B2", "#009E73"), labels = c("Scenedesmus","Chlamydomonas"))+
  xlab("Days")+
  ylab("Number of Cells *10^4 per ml")+
  theme_light()
```


# Sensitivity Analysis

Below I create functions to do the sensitivity analysis. First I made a function that takes for inputs the data, the optimal parameter esitmates, by what percent do you want to change the parameters, initial conditions, time, and the function that describes the differential equations and returns the % change in MSE from the optimal parameter estimates. 

```{r}
sensit=function(data, optimal_params, change, yini, t, fun){
  out_optimal=ode(yini, t, fun, optimal_params) # solve with optimal parameters
  out_optimal=out_optimal%>% #change the shape of the data for optimal parameters
    as.matrix()%>%
    as_tibble()%>%
    gather("a", "b", 2:dim(out_optimal)[2])
  data=data%>% # change shape for our current data
    gather("a", "b", 1:2)
  mse_optimal=mean((data$b-out_optimal$b)^2) # calculate current 
  mse_all=map_dbl(1:length(optimal_params), function(i){ # use a map since its faster than a for loop (essentially the same though)
    curr_param=optimal_params[i] # select parameter i
    param_set=optimal_params
    param_set[param_set==curr_param]=curr_param+curr_param*(change/100) # change parameter esitmate by a "change"" percent
    out_change=ode(yini, t, fun, param_set)# solve with new parameter estimates
    out_change=out_change%>% # reshape solved data
      as.matrix()%>%
      as_tibble()%>%
      gather("a", "b", 2:dim(out_optimal)[2])
      mse_change=mean((data$b-out_change$b)^2)# calculate MSE for changed params
      mse=abs(mse_change-mse_optimal)/mse_optimal # calculate percent change from optimal params
      return(mse)
  })
  return(mse_all)
}


```

Here is an example using the 1.7mM diluted and combined. As you can see the growth rate for chlamydomonas is the most sensitive at 0.245

```{r}

sensit(algae_dilluted_data_17dc, fitcomp_17dc$par, 1, c(.1,.1), algae_dilluted_data_17dc$day, fun=compderivs)

```


Next we built two functions that return the plotting data. We had to make two functions because the modFit function returns an object of type deSolve which doesnt coerce to a tibble very nicely(the order of columns was changing for different datasets), so this was the quickest way we could think of to solve the problem we were having. 

The function below is exactly the same just we return the data instead of calculating the % change in MSE.

```{r}
sensit_plots1=function(optimal_params, change, yini, t, fun){
  out_optimal=ode(yini, t, fun, optimal_params)
  out_optimal=out_optimal%>%
    as.matrix()%>%
    as_tibble()%>%
    rename("day"="time",
             "Scenedesmus"=`1`,
             "Chlamydomonas"=`2`)%>%
    gather("Species", "Count", 2:dim(out_optimal)[2])
  out_optimal$param_changed=0
  plotting_data=map_df(1:length(optimal_params), function(i){
    curr_param=optimal_params[i]
    param_set=optimal_params
    param_set[param_set==curr_param]=curr_param+curr_param*(change/100)
    out_change=ode(yini, t, fun, param_set)
    out_change1=out_change%>%
      as.matrix()%>%
      as_tibble()%>%
      rename("day"="time",
             "Scenedesmus"=`1`,
             "Chlamydomonas"=`2`)%>%
      gather("Species", "Count", 2:dim(out_change)[2])
      out_change1$param_changed=i
      return(out_change1)
  })
  plot_data=as_tibble(rbind(plotting_data, out_optimal))
  return(plot_data)
}
```


```{r}
sensit_plots=function(optimal_params, change, yini, t, fun){
  out_optimal=ode(yini, t, fun, optimal_params)
  out_optimal=out_optimal%>%
    as.matrix()%>%
    as_tibble()%>%
    rename("day"="time",
             "Scenedesmus"=`2`,
             "Chlamydomonas"=`1`)%>%
    gather("Species", "Count", 2:dim(out_optimal)[2])
  out_optimal$param_changed=0
  plotting_data=map_df(1:length(optimal_params), function(i){
    curr_param=optimal_params[i]
    param_set=optimal_params
    param_set[param_set==curr_param]=curr_param+curr_param*(change/100)
    out_change=ode(yini, t, fun, param_set)
    out_change1=out_change%>%
      as.matrix()%>%
      as_tibble()%>%
      rename("day"="time",
             "Scenedesmus"=`2`,
             "Chlamydomonas"=`1`)%>%
      gather("Species", "Count", 2:dim(out_change)[2])
      out_change1$param_changed=i
      return(out_change1)
  })
  plot_data=as_tibble(rbind(plotting_data, out_optimal))
  return(plot_data)
}
```


Below we make the plot that is seen in our poster

```{r}

plotting_dat=sensit_plots1(fitcomp_17dc$par, 10, c(.1,.1), seq(1,19,length=1000),  fun=compderivs)%>% # here we evaluate our functions on the three diluted and combined datasets. 
  rbind(sensit_plots(fitcomp_34dc$par, 10, c(.1,.1), seq(1,19,length=1000), fun=compderivs))%>%
  rbind(sensit_plots(fitcomp_5dc$par, 10, c(.1,.1), seq(1,19,length=1000), fun=compderivs))%>%
  mutate(id=1:42000)%>% # next we reshape everything and add some better lables. 
  mutate(Flask=ifelse(id<=14000,"1.7mM Dilluted and Combined",NA))%>%
  mutate(Flask=ifelse(id>14000 & id<=28000,"0.34mM Dilluted and Combined", Flask))%>%
  mutate(Flask=ifelse(id>2*42000/3,"5uM Dilluted and Combined", Flask))%>%
  mutate(Flask=fct_relevel(Flask, "5uM Dilluted and Combined"))


plt_dat_p1=plotting_dat%>% # selecting the optimal param portion of the data
  filter(param_changed==0)
plt_dat=plotting_dat%>% # selecting the optimal param portion of the data
  filter(param_changed!=0)


ggplot()+ # plotting the sensitivity analysis
  geom_line(aes(x=day, y=Count, color=as.factor(param_changed), linetype=Species), data=plt_dat_p1, size=2)+
  geom_line(aes(x=day, y=Count, color=as.factor(param_changed), linetype=Species), data=plt_dat, size=2, alpha=0.4)+
    ylab("Average Count (x10^4 cells/ml)")+
  xlab("Days")+
  theme_minimal()+
    scale_color_manual(values=c("#009E73", "#0072B2", "#999999", "#E69F00", "#56B4E9",  "#F0E442",  "#D55E00"), name="Parameter Changed", breaks=c("0", "1", "2", "3", "4", "5", "6"), labels=c("Optimal Values", "r_S", "K_S","a_S", "r_C", "K_C", "a_C"))+
  facet_wrap(~Flask)+
  theme(text = element_text(size = 20),
        axis.line = element_line(color="black"))+
  theme(plot.margin=unit(c(1,1,1.5,1.2),"cm"))+
  theme(legend.text=element_text(size=8),
        legend.title = element_text(size=12))
```



